MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-09-15 chapters/MetaTracts.tex
updated text

The directed Hausdorff distance is very sensitive to outliers in the data.
However, because MetaTracts after orientation clustering are constructed using cylinders with similar orientations, they are not plagued by outliers.
To cluster based on MetaTract proximity, we used single linkage hierarchical clustering.
Hierarchical clustering has a single parameter $h$, the desired number of clusters.
In hierarchical clustering each objects starts in its own cluster and clusters are merged based on some criterion.
Clusters are merged until there are only $h$ clusters left.
Hierarchical clustering is intuitive since it is easy to trace how clusters are formed and merged.
Single linkage clustering finds pairs of objects $p \in P$ and $q \in Q$ where $P \neq Q$ which are closer than other such pairs, and merges the containing clusters $P$ and $Q$.
We found that single linkage hierarchical clustering had two major drawbacks.
First, the clustering would produce some small clusters of just a few MetaTracts.
These MetaTracts were anomalies caused by noise in the data and did not represent true fiber bundles.
Second, if two fiber bundles were parallel for some of their length and then separated, they would sometimes be clustered into the same fiber bundle.
A short MetaTract which was parallel to both and did not extend into the separation region could form a link between the two fiber bundles, causing them to be clustered into a single bundle.
To address the problem of small clusters, we applied hierarchical clustering and then identified small clusters with few MetaTracts.
We removed the MetaTracts that were in those clusters from the data set and reapplied hierarchical clustering.
To address the problem of short MetaTracts joining different fiber bundles, we applied hierarchical clustering and then removed the
shortest fibers (fibers less than 0.6 times the median length) in each bundle. We then reapplied hierarchical clustering.
We repeated both until a steady state of clusters is reached and no new small fibers can be removed. The results of clustering the orientation clusters Fig.~\ref{fig:orientation_clustering}(A,C) are shown in Fig.~\ref{fig:orientation_clustering}(D,E) respectively. Note how the purple cluster Fig.~\ref{fig:orientation_clustering}(E) bundle ``7'' is extracted well.
\begin{figure}
\centering
		%\includegraphics[width=0.4\textwidth]{imagesMT2014/garb1}~
		\includegraphics[width=0.5\textwidth]{imagesMT2014/vol}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(A) Orientation wise decomposition of data set Figure~\ref{fig:data-char}A into the extracted bundles and (B) all of them combined.(C) Shows the results of voxelization of (B). (D) shows a 2D slice of the volume. (E,F) shows the extracted meshes.}
	\label{fig:crop-16-decomp}
\end{figure}

@@

Apart from direct visualization of the MetaTracts, we show two simple but yet highly important extensions for our domain specialist. The first one is to voxelize the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods are required for further analysis of the data, e.g. in simulations.
To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood and Figure~\ref{fig:crop-16-decomp}C and Figure~\ref{fig:prepreg}D shows the result of voxelization. Figure~\ref{fig:crop-16-decomp}(E, F) show examples of extracted meshes.

@@

 %glass
We tested our method on multiple datasets with varying characteristics. 
Figure~\ref{fig:crop-16-decomp} shows the result of data set~\ref{fig:data-char} decomposed into 2 orientation and each orientation decomposed into 10 clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 is 200 while the maximum MetaTract in the bundle is about 500 (measured in units of grid cube length). Fig.~\ref{fig:length_distribution}(B) shows the length per MetaTract distribution for a particular fiber bundle. 
Fig.~\ref{fig:prepreg} shows the results on our dataset2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300x350x300 and the datatype is uint16. The bundles are indistinguishable in the original data set Fig.~\ref{fig:prepreg}(A). A green dotted line shows one of the fiber bundle boundary. Fig.~\ref{fig:prepreg}(B) shows the fiber bundles for the two orientation. Fig.~\ref{fig:prepreg}(C) shows the combined results.Fig.~\ref{fig:prepreg}(D) shows the volume after voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done quickly in a few minutes on a Intel Xeon E5-2667 workstation according to user input `h'(section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without extra computation.

\section {User Evaluation}
\label {sec:evaluation}
%\begin{figure}[h]
%\centering
%\subfloat[]{\includegraphics[width=0.25\textwidth]{imagesMT2014/crop-6-F.PNG}}
%\subfloat[]{\includegraphics[width=0.25\textwidth]{imagesMT2014/orientB.PNG}}
%\caption{Contextual information: (a) Shows the four largest clusters in the data set in Fig~\ref{fig:prepreg}.
%(b) shows how the red bundle `bends' under the green bundle, which is not immediately apparent from just the volume rendering.}
%\label{fig:context}
%\end{figure}

\section{Parameter choices}
\label{sec:param_choices}
\begin{figure}
\centering
	\includegraphics[width=0.4\textwidth]{imagesMT2014/crop-16/histogram_10_A.PNG}
	\caption{Number of tracts and median, minimum and maximum length of individual MetaTracts in the orientation cluster Fig.~\ref{fig:orientation_clustering}, clustered into 10 clusters (Fig.~\ref{fig:crop-16-decomp}A top). The unit for length is the grid cube length. Figure~\ref{fig:length_distribution}(B) shows the length of individual MetaTracts in cluster 6. Clusters labeled 2,3,4,5 and 7 all have cardinality less than 10. }
	\label{fig:len_dist_crop16}
\end{figure}
The critical parameters are $K$ for the K-means in orientation based clustering and $h$ for the hierarchical clustering. K denotes the number of major directions, which is either known a priori or can be easily estimated by looking at the weaving pattern. 
Our framework is robust to the choice of $h$ (Section~\ref{sec:results}). Large $h$ keeps the major bundles intact. For example, in  Fig.~\ref{fig:len_dist_crop16}, where $h$ was set to ten while the ground truth was five, we observe that the major clusters remain well segmented and small clusters can be removed if they have too few elements (clusters 2,3,4 and 5 in Fig.~\ref{fig:len_dist_crop16} all have less than 10 elements). This is an appealing trait of the proximity based clustering, thus providing good results even when exact $h$ might be unknown.
The following parameters were fixed for all the tests. We set the reliable Hessian threshold $R_{h}$ to be 0.3 for all our tests. A $R_{h}$ of 0.0 would mean all points have reliable local orientation which would cause spurious MetaTracts detection. On the other hand a very high $ R_{h}$ would lead to be a decline in number of MetaTracts produced. The $\alpha$ and $\beta$ in $R_{h}$ are as explained in Frangi et al. \cite{Frangi1998} and set to 0.5. The length and the width parameters for the cylinders for the MetaTracts decide how coarse the fiber cylinders are. Larger cylinders will handle noisy local orientation better as it inspects a higher number of candidate points to extend the fiber. We used 10.00 and 2.00 for length and breadth, respectively for all tests. $\alpha$ and $\beta$ in equation~\ref{eqn:algo_1} decide how quickly the value of the factor decays, we have used [7-10] and half the length of cylinder (5.0). Our number of fiber bundle directions are limited, thus even for small $\eta$, the distinction between the orientation clusters is preserved quite well. We compared  $\eta=3-7$ experimentally without any dramatic change in results. 
%\begin{figure}
%\centering
% \subfloat[$\eta=7$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim3k2}}
%  \subfloat[$\eta=3$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim7k2}}
%  \caption{Embedding from $\mathbb{R}^m\to \mathbb{R}^\eta$, here m is the number of MetaTracts, 580 for this example. Followed by clustering into K=2 clusters.}
%  \label{fig:eta}
%\end{figure}
%\begin{figure}[htp]
%\centering
% \includegraphics[width=\linewidth]{imagesMetaTracts/kul-results.eps}
%  \caption{Results from orientation based dimensionality reduction and clustering. A is the original data set, B shows the results of fiber tracking. C shows the results of orientation based clustering and D shows the results of the proximity based clustering. }\label{fig:kull-full}
%\end{figure}

>>>>

The directed Hausdorff distance is very sensitive to outliers in the data.
However, because MetaTracts after orientation clustering are constructed using cylinders with similar orientations, they are not plagued by outliers.
To cluster based on MetaTract proximity, we used single linkage hierarchical clustering.
Hierarchical clustering has a single parameter $h$, the desired number of clusters.
In hierarchical clustering each objects starts in its own cluster and clusters are merged based on some criterion.
Clusters are merged until there are only $h$ clusters left.
Hierarchical clustering is intuitive since it is easy to trace how clusters are formed and merged.
Single linkage clustering finds pairs of objects $p \in P$ and $q \in Q$ where $P \neq Q$ which are closer than other such pairs, and merges the containing clusters $P$ and $Q$.
We found that single linkage hierarchical clustering had two major drawbacks.
First, the clustering would produce some small clusters of just a few MetaTracts.
These MetaTracts were anomalies caused by noise in the data and did not represent true fiber bundles.
Second, if two fiber bundles were parallel for some of their length and then separated, they would sometimes be clustered into the same fiber bundle.
A short MetaTract which was parallel to both and did not extend into the separation region could form a link between the two fiber bundles, causing them to be clustered into a single bundle.
To address the problem of small clusters, we applied hierarchical clustering and then identified small clusters with few MetaTracts.
We removed the MetaTracts that were in those clusters from the data set and reapplied hierarchical clustering.
To address the problem of short MetaTracts joining different fiber bundles, we applied hierarchical clustering and then removed the
shortest fibers (fibers less than 0.6 times the median length) in each bundle. We then reapplied hierarchical clustering.
We repeated both until a steady state of clusters is reached and no new small fibers can be removed. The results of clustering the orientation clusters Fig.~\ref{fig:orientation_clustering}(A,C) are shown in Fig.~\ref{fig:orientation_clustering}(D,E) respectively. Note how the purple cluster Fig.~\ref{fig:orientation_clustering}(E) bundle ``7'' is extracted well.
\begin{figure}
\centering
		%\includegraphics[width=0.4\textwidth]{imagesMT2014/garb1}~
		\includegraphics[width=0.5\textwidth]{imagesMT2014/image7b}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(A) Volume rendering of data set Figure~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
	\label{fig:crop-16-decomp}
\end{figure}

@@

Apart from direct visualization of the MetaTracts, we show two simple yet highly important extensions for our domain specialist. The first one is to voxelize the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods can possibly be used for further analysis of the data, e.g. in simulations.
To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood. Figure~\ref{fig:crop-16-decomp}A and Figure~\ref{fig:prepreg}D shows the result of voxelization. Figure~\ref{fig:crop-16-decomp}(C, D) show examples of extracted meshes.

@@

 %glass
We tested our method on multiple datasets with varying characteristics. 
Figure~\ref{fig:orientation_clustering} shows the result of data set~\ref{fig:data-char} decomposed into 2 orientation and each orientation decomposed into 10 clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 Fig.~\ref{fig:length_distribution} is 200 while the maximum MetaTract in the bundle is about 500 (measured in units of grid cube length). Fig.~\ref{fig:length_distribution}(B) shows the length per MetaTract distribution for a particular fiber bundle. 
Fig.~\ref{fig:prepreg} shows the results on our dataset2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300x350x300 and the datatype is uint16 (compared to dataset1 which was only uint8). The bundles are indistinguishable in the original data set Fig.~\ref{fig:prepreg}(A). A green dotted line shows one of the fiber bundle boundary. Fig.~\ref{fig:prepreg}(B) shows the fiber bundles for the two orientation. Fig.~\ref{fig:prepreg}(C) shows the combined results.Fig.~\ref{fig:prepreg}(D) shows the volume after voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done quickly in a few minutes on a Intel Xeon E5-2667 workstation according to user input `h'(section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without extra computation.

\section {User Evaluation}
\label {sec:evaluation}
We integrated MetaTracts into a tool and created a user evaluation questionnaire which we distributed.
%\begin{figure}[h]
%\centering
%\subfloat[]{\includegraphics[width=0.25\textwidth]{imagesMT2014/crop-6-F.PNG}}
%\subfloat[]{\includegraphics[width=0.25\textwidth]{imagesMT2014/orientB.PNG}}
%\caption{Contextual information: (a) Shows the four largest clusters in the data set in Fig~\ref{fig:prepreg}.
%(b) shows how the red bundle `bends' under the green bundle, which is not immediately apparent from just the volume rendering.}
%\label{fig:context}
%\end{figure}

\section{Parameter choices}
\label{sec:param_choices}
\begin{figure}
\centering
	\includegraphics[width=0.4\textwidth]{imagesMT2014/crop-16/histogram_10_A.PNG}
	\caption{Number of tracts and median, minimum and maximum length of individual MetaTracts in the orientation cluster Fig.~\ref{fig:orientation_clustering}, clustered into 10 clusters (Fig.~\ref{fig:crop-16-decomp}A top). The unit for length is the grid cube length. Figure~\ref{fig:length_distribution}(B) shows the length of individual MetaTracts in cluster 6. Clusters labeled 2,3,4,5 and 7 all have cardinality less than 10. }
	\label{fig:len_dist_crop16}
\end{figure}
The critical parameters are $K$ for the K-means in orientation based clustering and $h$ for the hierarchical clustering. K denotes the number of major directions, which is either known a priori or can be easily estimated by looking at the weaving pattern. 
Our framework is robust to the choice of $h$ (Section~\ref{sec:results}). Large $h$ keeps the major bundles intact. For example, in  Fig.~\ref{fig:len_dist_crop16}, where $h$ was set to ten while the ground truth was five, we observe that the major clusters remain well segmented and small clusters can be removed if they have too few elements (clusters 2,3,4,5 and 7 in Fig.~\ref{fig:len_dist_crop16} all have less than 10 elements). This is an appealing trait of the proximity based clustering, thus providing good results even when exact $h$ might be unknown.
The following parameters were fixed for all the tests. We set the reliable Hessian threshold $R_{h}$ to be 0.3 for all our tests. A $R_{h}$ of 0.0 would mean all points have reliable local orientation which would cause spurious MetaTracts detection. On the other hand a very high $ R_{h}$ would lead to be a decline in number of MetaTracts produced. The $\alpha$ and $\beta$ in $R_{h}$ are as explained in Frangi et al. \cite{Frangi1998} and set to 0.5. The length and the width parameters for the cylinders for the MetaTracts decide how coarse the fiber cylinders are. Larger cylinders will handle noisy local orientation better as it inspects a higher number of candidate points to extend the fiber. We used 10.00 and 2.00 for length and breadth, respectively for all tests. $\alpha$ and $\beta$ in equation~\ref{eqn:algo_1} decide how quickly the value of the factor decays, we have used [7-10] and half the length of cylinder (5.0). Our number of fiber bundle directions are limited, thus even for small $\eta$, the distinction between the orientation clusters is preserved quite well. We compared  $\eta=3-7$ experimentally without any dramatic change in results. 
%\begin{figure}
%\centering
% \subfloat[$\eta=7$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim3k2}}
%  \subfloat[$\eta=3$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim7k2}}
%  \caption{Embedding from $\mathbb{R}^m\to \mathbb{R}^\eta$, here m is the number of MetaTracts, 580 for this example. Followed by clustering into K=2 clusters.}
%  \label{fig:eta}
%\end{figure}
%\begin{figure}[htp]
%\centering
% \includegraphics[width=\linewidth]{imagesMetaTracts/kul-results.eps}
%  \caption{Results from orientation based dimensionality reduction and clustering. A is the original data set, B shows the results of fiber tracking. C shows the results of orientation based clustering and D shows the results of the proximity based clustering. }\label{fig:kull-full}
%\end{figure}
